Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-08 paper/main.tex
Updates
Improve abstract

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative method that guides
researchers in manually constructing or improving the fit of an existing DAG
while incorporating their expert knowledge. Compared to other expert knowledge
specification methods, this method explicitly asks for the causal ordering
between two specific variables in each step. Additionally, we provide a way to
rank the potential new edge addition that improves the model fit the most in
each step. Empirical results show that this guided manual model construction
approach achieves performance comparable to automated algorithms if we are able
to correctly decide the orientation of new edges in one out of three cases.
Additionally, we demonstrate that, in the absence of an expert, a Large
Language Model (LLM) can be potentially used to determine edge orientations. We
provide the implementation of our approach in both a web tool at: <redacted for
review> and a Python package <redacted for review>.

@@

@@

To enable users to apply this method to their own datasets, we developed an
interactive web-based tool (shown in Fig.~\ref{fig:web}) for constructing
models. Users begin by uploading their dataset, which initializes an empty DAG
with nodes corresponding to the dataset\'s variables. They can then specify a
p-value threshold and a measure of association threshold.

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on domain knowledge. This preference arises due to several practical
challenges with automated algorithms, such as their tendency to make obvious
errors and output Markov Equivalence Classes (MECs) rather than a DAG. To
address these challenges, we propose an iterative DAG construction that
combines domain knowledge with data-driven insights. Our approach iteratively
identifies pair of variables with a missing edge in the DAG, and asks for the
orientation of the edge from an expert. We also propose a method to rank these
potential new edges based on their impact on the model fit. Empirical results
show that this iterative model construction approach achieves performance
comparable to automated algorithms if we are able to correctly decide the
orientation of new edges in one out of three cases. Additionally, we
demonstrate that, in the absence of an expert, a Large Language Model (LLM) can
be used to determine edge orientations. We provide the implementation of our
approach in both a web tool at: <redacted for review> and a Python package
<redacted for review>.

@@

Given a DAG $ G = (V, E) $, the total unexplained correlation, $\tau$ in define as:
\begin{equation}
	\tau = \sum_{\{ (X, Y) \in V | X \rightarrow Y \not \in E \wedge Y \rightarrow X \not \in E \} } \phi(X, Y, \mathrm{pa}_G(X) \cup \mathrm{pa}_G(Y)))
\end{equation}

@@

To enable users to apply this method to their own datasets, we developed an
interactive web-based tool (shown in Fig.~\ref{fig:web}) for constructing
models. Users begin by uploading their dataset, which initializes an empty DAG
with nodes corresponding to the dataset\'s variables. They can then specify a
p-value threshold and a measure of association threshold. We use the measure of 
conditional association shown in Appendix~\ref{sec:mixed_association}.
